---
title: "Bayesian Estimation of First-Year Retention Rates: A Comparative Analysis of Commuter and Hispanic-Serving Institutions in Texas"
author: "Eugene, Shikyna, Rehana"
date: "`r Sys.Date()`"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    code-fold: true
    theme: flatly
    highlight-style: tango
    include-in-header:
      - text: |
          <style>
          .quarto-title-block .quarto-title-banner,
          .quarto-title-block .quarto-title {
            text-align: center;
          }
          </style>
execute:
  warning: false
  message: false
  echo: true
editor: 
  markdown: 
    wrap: 72
---

## Setup & data preparation

```{r, include=FALSE}
library(tidyverse)
library(rjags)
library(coda)
library(bayesplot)
library(knitr)
library(kableExtra)
library(plotly)
```

```{r, message=FALSE, warning=FALSE}
y22 <- read_csv("Bayes CommuterHSI Data/2223.csv")
y23 <- read_csv("Bayes CommuterHSI Data/2324.csv")
y24 <- read_csv("Bayes CommuterHSI Data/2425.csv")
hsi22 <- read_csv("Bayes CommuterHSI Data/HSI22.csv")
hsi23 <- read_csv("Bayes CommuterHSI Data/HSI23.csv")
hsi24 <- read_csv("Bayes CommuterHSI Data/HSI24.csv")

y22 <- y22 %>%
  select(-starts_with("CUSTOMCGIDS"), -'institution name...4', -'HD2022.State abbreviation', -'IC2022_PCCAMPUSES.Sector of institution') %>% distinct()
y23 <- y23 %>%
  select(-starts_with("CUSTOMCGIDS"), -'institution name...4', -'HD2023.State abbreviation', -'IC2023_PCCAMPUSES.Sector of institution') %>% distinct()
y24 <- y24 %>%
  select(-starts_with("CUSTOMCGIDS"), -'institution name...4', -'HD2024.State abbreviation', -'XROOMCAP', -'XRET_PCF', -'XRET_NMF', -'XRRFTCTA', -'HD2024.Sector of institution') %>% distinct()

hsi22 <- hsi22 %>% 
  select(-'IDX_EF', -'EF2022A.Level of student')
hsi23 <- hsi23 %>% 
  select(-'IDX_EF', -'EF2023A.Level of student')
hsi24 <- hsi24 %>% 
  select(-'IDX_EF', -'EF2024A.Level of student') %>% rename('hisp_pop' = 'EF2024A.Hispanic total', 'tot_pop' = 'EF2024A.Grand total')

y22 <- y22 %>% rename('name' = 'institution name...2', 'housing' = 'IC2022.Institution provide on-campus housing', 'dorm_cap' = 'IC2022.Total dormitory capacity', 'cohort21' = 'EF2022D_RV.Full-time adjusted fall 2021 cohort', 'return_cohort22' = 'EF2022D_RV.Students from the full-time adjusted fall 2021 cohort enrolled in fall 2022', 'ret_rate22' = 'EF2022D_RV.Full-time retention rate, 2022')

y23 <- y23 %>% rename('name' = 'institution name...2', 'housing' = 'IC2023.Institution provide nstitutionally-controlled housing (on-campus and/or off-campus)', 'dorm_cap' = 'IC2023.Housing capacity', 'cohort22' = 'EF2023D_RV.Full-time adjusted fall 2022 cohort', 'return_cohort23' = 'EF2023D_RV.Students from the full-time adjusted fall 2022 cohort enrolled in fall 2023', 'ret_rate23' = 'EF2023D_RV.Full-time retention rate, 2023')

y24 <- y24 %>% rename('name' = 'institution name...2', 'housing' = 'COST1_2024.Institution provide Institutionally-controlled housing (on-campus and/or off-campus)', 'dorm_cap' = 'COST1_2024.Housing capacity', 'cohort23' = 'EF2024D.Full-time adjusted fall 2023 cohort', 'return_cohort24' = 'EF2024D.Students from the full-time adjusted fall 2023 cohort enrolled in fall 2024', 'ret_rate24' = 'EF2024D.Full-time retention rate, 2024') %>% filter(cohort23 > 0)

# Classify institutions: HSI & Commuter Status
y24 <- left_join(y24, select(hsi24, unitid, tot_pop, hisp_pop), by='unitid') %>% mutate('hsi' = (hisp_pop / tot_pop) > .25, 'commuter' = (dorm_cap / tot_pop) < .25)
y24$commuter[y24$housing == "No"] <- T

y24 <- y24 %>% select(-'housing', -'dorm_cap', -'hisp_pop', -'year')

uni <- y24 %>% left_join(select(y23, 'unitid', 'cohort22', 'return_cohort23', 'ret_rate23'), by='unitid') %>% left_join(select(y22, 'unitid', 'cohort21', 'return_cohort22', 'ret_rate22'), by='unitid')

uni <- uni %>% select("unitid", "name", "tot_pop", "hsi", "commuter", "cohort23", "return_cohort24", "ret_rate24", "cohort22", "return_cohort23", "ret_rate23", "cohort21", "return_cohort22", "ret_rate22")
uni <- uni %>% filter(unitid != "229355")
```

In the final table, we have `r ncol(uni)` variables. This includes; the
school's ID, name, population, commuter and HSI flags, and the cohort
and retention rate information for the years 2022, 2023, and 2024.

## Descriptive Summary

```{r}
uni %>% select(tot_pop, hsi, commuter, ret_rate24, ret_rate23, ret_rate22) %>% summarise(
  n_schools = nrow(uni),
  hsi_proportion = round(sum(uni$hsi) / nrow(uni),  2),
  commuter_proportion = round(sum(uni$commuter) / nrow(uni), 2),
  avg_ret_rate24 = mean(ret_rate24),
  avg_ret_rate23 = mean(ret_rate23),
  avg_ret_rate22 = mean(ret_rate22),
) %>% pivot_longer(everything(), names_to = "Statistic", values_to = "Value") %>%
  kable(digits = 3, caption = "Summary of Texas public 4-year institutions") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    position = "center"
  )
```

# Part 1: Conjugate Beta-Binomial Analysis (base model)

The baseline model pools all students within each of the four groups

$$k_g \mid n_g, \theta_g \sim \text{Binomial}(n_g, \theta_g)$$

$$\theta_g \sim \text{Beta}(\alpha_0, \beta_0)$$

The conjugate posterior would be:

$$\theta_g \mid k_g \sim \text{Beta}(\alpha_0 + k_g, \beta_0 + n_g - k_g)$$

**From Lectures: 7, 9**

## Pool the data into groups

We would group each and find retained cohort by year using summarize.

## Test three prior choices for sensitivity analysis

We can test;

Uninformative: Beta (1, 1) - Completely flat prior

```{r, echo=F}
plot_ly(x = seq(0, 1, length.out = 100), y = dbeta(seq(0, 1, length.out = 100), 1, 1), type = 'scatter', mode = 'lines', fill = 'tozeroy') %>%
  layout(
    title = paste0("Beta Distribution (Alpha =", 1, ", Beta =", 1, ")"),
    xaxis = list(title = "x"),
    yaxis = list(title = "Density")
  )
```

\
Weakly Informative: Beta (2, 2) - Slight pull to 0.5

```{r, echo=F}
plot_ly(x = seq(0, 1, length.out = 100), y = dbeta(seq(0, 1, length.out = 100), 2, 2), type = 'scatter', mode = 'lines', fill = 'tozeroy') %>%
  layout(
    title = paste0("Beta Distribution (Alpha =", 2, ", Beta =", 2, ")"),
    xaxis = list(title = "x"),
    yaxis = list(title = "Density")
  )
```

Informative: Beta (7, 3) - Centered near our calculated 71.5% average
retention rate.

```{r, echo=F}
plot_ly(x = seq(0, 1, length.out = 100), y = dbeta(seq(0, 1, length.out = 100), 7, 3), type = 'scatter', mode = 'lines', fill = 'tozeroy') %>%
  layout(
    title = paste0("Beta Distribution (Alpha =", 7, ", Beta =", 3, ")"),
    xaxis = list(title = "x"),
    yaxis = list(title = "Density")
  )
```

\*Make sure to include credible intervals

-- Maybe consider using posterior of previous years as prior for
following years.

-- Use posterior predictive to check adequacy of our model.

**From lectures: 4, 9, 10**

## Compute posteriors under each prior

## Plot posterior densities (main prior)

## Posterior probability statements

Such as probability HSI-Commuter group has lower retention than
residential groups

# Part 2: Hierarchical Beta-Binomial Model (Main Analysis)

Because retention is a count of successes out of a fixed cohort, we use
the Binomial likelihood with a conjugate Beta prior rather than the
Normal-Normal hierarchy from class (l18).

The conjugate model we used assumes all the institutions within a group
share the same retention probability, which is kind of unrealistic. The
hierarchical model allows each institution (i) have its own (theta_i),
but institutions within the same group (g) share a common Beta
distribution:

$$\text{Layer 1 (likelihood):} \quad k_i \mid n_i, \theta_i \sim \text{Binomial}(n_i, \theta_i)$$

$$\text{Layer 2 (prior):} \quad \theta_i \mid \alpha_{g[i]}, \beta_{g[i]} \sim \text{Beta}(\alpha_{g[i]}, \beta_{g[i]})$$

$$\text{Layer 3 (hyperprior):} \quad \mu_g \sim \text{Beta}(8, 2), \quad \kappa_g \sim \text{Gamma}(2, 0.5) + 2$$

where $\alpha_g = \mu_g \kappa_g$ and $\beta_g = (1 - \mu_g) \kappa_g$.
This reparameterization makes $\mu_g$ the group mean and $\kappa_g$ a
concentration parameter (larger = tighter around the group mean).

We chose the nation average for the Beta prior with concentration around
80%.

We chose Gamma with a mean of 4 to allow kappa to range broadly. I also
seen that sometimes it is recommended to add a constant, because Beta
distributions with a value less than 1 are U-shaped, which wouldn't make
sense with retention rates.

**From lectures: 17, 18, 19**

## Prepare data for JAGS

## JAGS model

**From lectures: 8, 16**

## Compile and fit

```{r}
# model <- 

# update()

# posterior_sample
```

## If we're doing MCMC, we do that now

**From lectures: 16**

-- Create MCMC

## Group-level posterior summaries

\*Make sure to include credible intervals for each group's
hyperparameter mean using quantile(mu_samples, c(0.025, 0.975))

-- Plot & compare credible intervals by year for each group

## Institution-level posteriors / shrinkage

\*Make sure to include credible intervals. This is where the smaller
schools will have their intervals pulled toward the group mean more than
larger schools.

## Posterior probability statements from the hierarchical model

# Part 3: Model Comparison

Compare conjugate pooled posteriors against hierarchical group-level
posteriors for the same four groups. The hierarchical intervals will be
wider for small groups (appropriate acknowledgment of institutional
heterogeneity) and may shift means slightly toward the overall data
pattern.

\*Make sure to include credible intervals from conjugate and
hierarchical models

-- Maybe try a multinomial logistic regression to compare all the groups
adds rigor, rather than doing all the groups independently.

## Bayesian Anova

**From lectures: 17, 19, 20**

# Conclusion:

Points to hit:

-   Key findings on HSI vs non-HSI and commuter vs residential retention
-   Specifically what the posteriors say about institutions like UHD
-   The value added by the hierarchical model (shrinkage, institutional
    heterogeneity)
-   Limitations: single year, institution-level aggregation ignores
    student-level predictors, etc.
-   Future directions: multi-year trend, student-level data, additional
    covariates

**From lectures: 5** - Has good language on interpretations

# References

IPEDS
