---
title: "Bayesian Estimation of First-Year Retention Rates: A Comparative Analysis of Commuter and Hispanic-Serving Institutions in Texas"
author:   
  - Eugene Kazi
  - Shikyna William
  - Rehana Hajee
date: "`r Sys.Date()`"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    code-fold: true
    theme: flatly
    highlight-style: tango
    include-in-header:
      - text: |
          <style>
          .quarto-title-block .quarto-title-banner,
          .quarto-title-block .quarto-title {
            text-align: center;
          }
          .quarto-title-meta > div {
            text-align: center;
          }
          </style>
  #pdf:
    #toc: true
    #number-sections: true
execute:
  warning: false
  message: false
  echo: true
editor: 
  markdown: 
    wrap: 72
---

# Introduction, Problem, Purpose

### Opening

First-year retention is one of the most watched indicators of
institutional performance in higher education. Placed in the heart of
one of the largest and most diverse cities in the United States, our
school, the University of Houston-Downtown carries the attribute of a
campus that lacks housing, and is proudly recognized as a Hispanic
Serving Institution. These characteristics of our campus raised the
question on whether any recent studies on the effect of retention rate
throughout Texas exists. With UHD being both an HSI and a commuter-heavy
campus, these two traits intersect at most Texas public 4-year
universities but are rarely studied together. This project applies
Bayesian methods to estimate first-year retention rates across Texas
public 4-year institutions, with a focus on whether HSI status and
commuter status interact to produce different retention patterns.

### Background & Related Studies

The literature on student retention has long emphasized institutional
context as a predictor of student outcomes. Astin (1993) and other
foundational researchers established a positive association between
living on campus and retention, finding that students with on-campus
residency tended to be retained at higher rates than commuters. However,
more recent work complicates this picture. A 2020 mixed-methods study at
a Midwest university found first-year to second-year retention rates of
63.21% for commuters and 66.07% for residential students, a difference
that was not statistically significant, suggesting the residential
advantage may be smaller than once thought, (Andriano 2020).

Research on Hispanic-Serving Institutions presents a similarly nuanced
picture. Retention rates at HSIs are slightly higher than national
averages for all students at institutions of higher education (67%
versus 62%), despite serving student populations that face structurally
higher barriers to completion, (Postsecondary National Policy Institute
2021). Recent research suggests that students at HSIs report a more
supportive environment and greater levels of satisfaction than peers at
predominantly white institutions, contributing to higher retention
through a stronger sense of belonging, (National Center for Education
Statistics).

What is missing from this literature is a Bayesian treatment of how HSI
status and commuter status combine. Most existing studies treat each
factor in isolation or as a frequentist comparison, which is poorly
suited to the small institution-level samples available within a single
state.

Based on the existing literature, we hypothesized that the HSI-Commuter
group would show the lowest retention rates, given that both HSI status
and commuter status have individually been associated with structural
challenges to student persistence. We further hypothesized that
NonHSI-Residential institutions would show the highest retention,
consistent with the literature on residential advantages and the
historical performance of well-resourced public flagships.

Our study focuses on three questions:

1.  What are the posterior distributions of first-year retention
    probability for each of the four institutional groups?
2.  Does the HSI-Commuter group, which describes UHD and similar
    institutions, show meaningfully different retention from the other
    three groups?
3.  Are these patterns stable across the three most recent cohort years,
    or do they vary substantially?

## Setup & data preparation

```{r, include=FALSE}
library(tidyverse)
library(rjags)
library(coda)
library(bayesplot)
library(knitr)
library(kableExtra)
library(plotly)
library(ggridges)
set.seed(2026)
```

```{r data_prep, message=FALSE, warning=FALSE}
y22 <- read_csv("Bayes CommuterHSI Data/2223.csv")
y23 <- read_csv("Bayes CommuterHSI Data/2324.csv")
y24 <- read_csv("Bayes CommuterHSI Data/2425.csv")
hsi22 <- read_csv("Bayes CommuterHSI Data/HSI22.csv")
hsi23 <- read_csv("Bayes CommuterHSI Data/HSI23.csv")
hsi24 <- read_csv("Bayes CommuterHSI Data/HSI24.csv")

y22 <- y22 %>%
  select(-starts_with("CUSTOMCGIDS"), -'institution name...4', -'HD2022.State abbreviation', -'IC2022_PCCAMPUSES.Sector of institution') %>% distinct()
y23 <- y23 %>%
  select(-starts_with("CUSTOMCGIDS"), -'institution name...4', -'HD2023.State abbreviation', -'IC2023_PCCAMPUSES.Sector of institution') %>% distinct()
y24 <- y24 %>%
  select(-starts_with("CUSTOMCGIDS"), -'institution name...4', -'HD2024.State abbreviation', -'XROOMCAP', -'XRET_PCF', -'XRET_NMF', -'XRRFTCTA', -'HD2024.Sector of institution') %>% distinct()

hsi22 <- hsi22 %>% 
  select(-'IDX_EF', -'EF2022A.Level of student')
hsi23 <- hsi23 %>% 
  select(-'IDX_EF', -'EF2023A.Level of student')
hsi24 <- hsi24 %>% 
  select(-'IDX_EF', -'EF2024A.Level of student') %>% rename('hisp_pop' = 'EF2024A.Hispanic total', 'tot_pop' = 'EF2024A.Grand total')

y22 <- y22 %>% rename('name' = 'institution name...2', 'housing' = 'IC2022.Institution provide on-campus housing', 'dorm_cap' = 'IC2022.Total dormitory capacity', 'cohort21' = 'EF2022D_RV.Full-time adjusted fall 2021 cohort', 'return_cohort22' = 'EF2022D_RV.Students from the full-time adjusted fall 2021 cohort enrolled in fall 2022', 'ret_rate22' = 'EF2022D_RV.Full-time retention rate, 2022')

y23 <- y23 %>% rename('name' = 'institution name...2', 'housing' = 'IC2023.Institution provide nstitutionally-controlled housing (on-campus and/or off-campus)', 'dorm_cap' = 'IC2023.Housing capacity', 'cohort22' = 'EF2023D_RV.Full-time adjusted fall 2022 cohort', 'return_cohort23' = 'EF2023D_RV.Students from the full-time adjusted fall 2022 cohort enrolled in fall 2023', 'ret_rate23' = 'EF2023D_RV.Full-time retention rate, 2023')

y24 <- y24 %>% rename('name' = 'institution name...2', 'housing' = 'COST1_2024.Institution provide Institutionally-controlled housing (on-campus and/or off-campus)', 'dorm_cap' = 'COST1_2024.Housing capacity', 'cohort23' = 'EF2024D.Full-time adjusted fall 2023 cohort', 'return_cohort24' = 'EF2024D.Students from the full-time adjusted fall 2023 cohort enrolled in fall 2024', 'ret_rate24' = 'EF2024D.Full-time retention rate, 2024') %>% filter(cohort23 > 0)

# Classify institutions: HSI & Commuter Status
y24 <- left_join(y24, select(hsi24, unitid, tot_pop, hisp_pop), by='unitid') %>% mutate('hsi' = (hisp_pop / tot_pop) > .25, 'commuter' = (dorm_cap / tot_pop) < .25)
y24$commuter[y24$housing == "No"] <- T

y24 <- y24 %>% select(-'housing', -'dorm_cap', -'hisp_pop', -'year')

uni <- y24 %>% left_join(select(y23, 'unitid', 'cohort22', 'return_cohort23', 'ret_rate23'), by='unitid') %>% left_join(select(y22, 'unitid', 'cohort21', 'return_cohort22', 'ret_rate22'), by='unitid')

uni <- uni %>% select("unitid", "name", "tot_pop", "hsi", "commuter", "cohort23", "return_cohort24", "ret_rate24", "cohort22", "return_cohort23", "ret_rate23", "cohort21", "return_cohort22", "ret_rate22")
uni <- uni %>% filter(unitid != "229355")

uni <- uni %>%
  mutate(
    group = case_when(
      hsi == TRUE  & commuter == TRUE  ~ "HSI_Commuter",
      hsi == TRUE  & commuter == FALSE ~ "HSI_Residential",
      hsi == FALSE & commuter == TRUE  ~ "NonHSI_Commuter",
      hsi == FALSE & commuter == FALSE ~ "NonHSI_Residential"
    ),
    group = factor(group, levels = c("HSI_Commuter", "HSI_Residential",
                                     "NonHSI_Commuter", "NonHSI_Residential")))
n_inst   <- nrow(uni)
n_groups <- length(levels(uni$group))
uni$group_idx <- as.integer(uni$group)
```

Our data come from the Integrated Postsecondary Education Data System
(IPEDS), maintained by the National Center for Education Statistics. We
extracted institution-level retention data for the 2021, 2022, and 2023
entering cohorts (measured for retention in fall 2022, 2023, and 2024
respectively). Our study is restricted to Texas public 4-year
institutions with positive entering cohort sizes, yielding 38
institutions. We classified each institution along two dimensions: HSI
status - institutions with at least 25% Hispanic undergraduate
enrollment, following the federal designation criterion, and Commuter
status - institutions with on-campus housing capacity below 25% of total
undergraduate enrollment, or with no on-campus housing offered.

## Descriptive Summary

```{r desc_summary}
uni %>% select(tot_pop, hsi, commuter, ret_rate24, ret_rate23, ret_rate22) %>% summarise(
  n_schools = nrow(uni),
  hsi_proportion = round(sum(uni$hsi) / nrow(uni),  2),
  commuter_proportion = round(sum(uni$commuter) / nrow(uni), 2),
  avg_ret_rate24 = mean(ret_rate24),
  avg_ret_rate23 = mean(ret_rate23),
  avg_ret_rate22 = mean(ret_rate22),
) %>% pivot_longer(everything(), names_to = "Statistic", values_to = "Value") %>%
  kbl(digits = 3, caption = "Summary of Texas public 4-year institutions") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    position = "center"
  )
```

Our final dataset includes 38 Texas public 4-year institutions. Of
these, 79% (30 schools) qualify as Hispanic-Serving Institutions under
the federal threshold of 25% Hispanic undergraduate enrollment, and 58%
(22 schools) qualify as commuter institutions under our threshold of
fewer than 25% on-campus housing capacity. The state-wide average
retention rate has been remarkably stable across the three cohorts
examined, ranging from 71.18% to 71.82%. This stability supports
treating each year as a comparable replication in our analyses.

# Part 1: Conjugate Beta-Binomial Analysis (base model)

The baseline model pools all students within each of the four groups:

$$X_g \mid n_g, \theta_g \sim \text{Binomial}(n_g, \theta_g)$$

```{r}
# Likelihood
likelihood_p1 <- data.frame(
  Symbol = c("\\(X_g\\)", "\\(n_g\\)", "\\(\\theta_g\\)", "\\(\\text{Bin}(n, \\theta)\\)"),
  Name = c("Pooled retained count (random variable)",
           "Pooled cohort size",
           "True retention probability for group \\(g\\)",
           "Binomial distribution"),
  Meaning = c("Total number of students across all institutions in group \\(g\\) who returned for their second year",
              "Total number of full-time first-time students across all institutions in group \\(g\\)",
              "Unknown true probability that a student in group \\(g\\) returns for their second year",
              "Out of \\(n\\) independent trials, each succeeding with probability \\(\\theta\\)"),
  Source = c("Observed data, summed within group",
             "Observed data, summed within group",
             "Unknown parameter, estimated analytically",
             "Distribution shape")
)

kbl(likelihood_p1, escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                latex_options = c("striped", "hold_position"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE, background = "#34495E", color = "white")
```

$$\theta_g \sim \text{Beta}(\alpha_0, \beta_0)$$

```{r}
# Prior
prior_p1 <- data.frame(
  Symbol = c("\\(\\theta_g\\)", "\\(\\alpha\\)", "\\(\\beta\\)", "\\(\\text{Beta}(\\alpha, \\beta)\\)"),
  Name = c("True retention probability for group \\(g\\)",
           "Prior shape parameter (alpha)",
           "Prior shape parameter (beta)",
           "Beta distribution"),
  Meaning = c("Same as in the likelihood",
              "Interpretable as the number of \"prior successes\"",
              "Interpretable as the number of \"prior failures\"",
              "Flexible distribution on \\([0, 1]\\) defined by two shape values"),
  Source = c("Unknown parameter",
             "Fixed by us (varies by prior choice: 1, 2, or 7)",
             "Fixed by us (varies by prior choice: 1, 2, or 3)",
             "Distribution shape")
)

kbl(prior_p1, escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                latex_options = c("striped", "hold_position"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE, background = "#34495E", color = "white")
```

The conjugate posterior would be:

$$\theta_g \mid x_g \sim \text{Beta}(\alpha_0 + x_g, \beta_0 + n_g - x_g)$$

```{r}
# Posterior
posterior_p1 <- data.frame(
  Symbol = c("\\(\\theta_g \\mid x_g\\)", "\\(\\alpha + x_g\\)", "\\(\\beta + n_g - x_g\\)"),
  Name = c("Posterior distribution of \\(\\theta_g\\) given the data",
           "Updated alpha (posterior shape parameter 1)",
           "Updated beta (posterior shape parameter 2)"),
  Meaning = c("Our updated belief about \\(\\theta_g\\) after observing \\(x_g\\) retained students out of \\(n_g\\)",
              "Prior alpha plus observed successes",
              "Prior beta plus observed failures"),
  Source = c("Conjugate result (closed-form Beta)",
             "Computed from data and prior",
             "Computed from data and prior")
)

kbl(posterior_p1, escape = FALSE, caption = "Posterior (conjugate update)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                latex_options = c("striped", "hold_position"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE, background = "#34495E", color = "white")
```

**From Lectures: 7, 9**

## Pool the data into groups

We would group each and find retained cohort by year using summarize.

## Test three prior choices for sensitivity analysis

We can test;

Uninformative: Beta (1, 1) - Completely flat prior

```{r, echo=F}
plot_ly(x = seq(0, 1, length.out = 100), y = dbeta(seq(0, 1, length.out = 100), 1, 1), type = 'scatter', mode = 'lines', fill = 'tozeroy') %>%
  layout(
    title = paste0("Beta Distribution (Alpha =", 1, ", Beta =", 1, ")"),
    xaxis = list(title = "x"),
    yaxis = list(title = "Density")
  )
```

\
Weakly Informative: Beta (2, 2) - Slight pull to 0.5

```{r, echo=F}
plot_ly(x = seq(0, 1, length.out = 100), y = dbeta(seq(0, 1, length.out = 100), 2, 2), type = 'scatter', mode = 'lines', fill = 'tozeroy') %>%
  layout(
    title = paste0("Beta Distribution (Alpha =", 2, ", Beta =", 2, ")"),
    xaxis = list(title = "x"),
    yaxis = list(title = "Density")
  )
```

Informative: Beta (7, 3) - Centered near our calculated 71.5% average
retention rate.

```{r, echo=F}
plot_ly(x = seq(0, 1, length.out = 100), y = dbeta(seq(0, 1, length.out = 100), 7, 3), type = 'scatter', mode = 'lines', fill = 'tozeroy') %>%
  layout(
    title = paste0("Beta Distribution (Alpha =", 7, ", Beta =", 3, ")"),
    xaxis = list(title = "x"),
    yaxis = list(title = "Density")
  )
```

\*Make sure to include credible intervals

-- Maybe consider using posterior of previous years as prior for
following years.

-- Use posterior predictive to check adequacy of our model.

**From lectures: 4, 9, 10**

## Compute posteriors under each prior

## Plot posterior densities (main prior)

## Posterior probability statements

Such as probability HSI-Commuter group has lower retention than
residential groups

# Part 2: Hierarchical Beta-Binomial Model (Main Analysis)

Our hierarchical model extends the conjugate framework by adding
institution-level retention probabilities θ_ig within each group. We
adapt the three-layer Normal-Normal hierarchy from Lecture 18 by
replacing the Normal likelihood with a Binomial and the Normal prior
with a Beta, since retention is a count of successes out of a fixed
cohort.

The conjugate model we used assumes all the institutions within a group
share the same retention probability, which is kind of unrealistic. The
hierarchical model allows each institution (i) have its own (theta_i),
but institutions within the same group (g) share a common Beta
distribution:

$$\text{Layer 1 (likelihood):} \quad X_{ig} \mid n_{ig}, \theta_{ig} \sim \text{Binomial}(n_{ig}, \theta_{ig})$$

```{r}
# Layer 1: Likelihood
layer1 <- data.frame(
  Symbol = c("\\(X_{ig}\\)", "\\(n_{ig}\\)", "\\(\\theta_{ig}\\)", "\\(\\text{Bin}(n, \\theta)\\)"),
  Name = c("Retained count (random variable)",
           "Cohort size",
           "True retention probability",
           "Binomial distribution"),
  Meaning = c("Number of students from institution \\(i\\) in group \\(g\\) who returned for their second year",
              "Total full-time first-time students at institution \\(i\\) in group \\(g\\) who entered the previous fall",
              "Unknown true probability that a student at institution \\(i\\) in group \\(g\\) returns for year two",
              "Out of \\(n\\) independent trials, each succeeding with probability \\(\\theta\\)"),
  Source = c("Observed data (`retained` column)",
             "Observed data (`cohort_n` column)",
             "Unknown parameter, estimated by JAGS",
             "Distribution shape")
)

kbl(layer1, escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                latex_options = c("striped", "hold_position"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE, background = "#34495E", color = "white")
```

$$\text{Layer 2 (prior):} \quad \theta_{ig} \mid \alpha_{g}, \beta_g \sim \text{Beta}(\alpha_g, \beta_g)$$

```{r}
# Layer 2: Prior
layer2 <- data.frame(
  Symbol = c("\\(\\theta_{ig}\\)", "\\(\\alpha_g\\)", "\\(\\beta_g\\)", "\\(\\text{Beta}(\\alpha, \\beta)\\)"),
  Name = c("True retention probability",
           "Beta shape parameter (alpha) for group \\(g\\)",
           "Beta shape parameter (beta) for group \\(g\\)",
           "Beta distribution"),
  Meaning = c("Same as Layer 1",
              "Prior successes parameter for the group-level Beta",
              "Prior failures parameter for the group-level Beta",
              "Flexible distribution on \\([0, 1]\\) defined by two shape values"),
  Source = c("Unknown parameter",
             "Computed: \\(\\alpha_g = \\mu_g \\cdot \\kappa_g\\)",
             "Computed: \\(\\beta_g = (1 - \\mu_g) \\cdot \\kappa_g\\)",
             "Distribution shape")
)

kbl(layer2, escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                latex_options = c("striped", "hold_position"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE, background = "#34495E", color = "white")
```

$$\text{Layer 3 (hyperprior):} \quad \mu_g \sim \text{Beta}(7, 3), \quad \kappa_g \sim \text{Gamma}(2, 0.5) + 2$$

```{r}
# Layer 3: Hyperprior
layer3 <- data.frame(
  Symbol = c("\\(\\mu_g\\)", "\\(\\kappa_g\\)", "\\(\\text{Beta}(7, 3)\\)", "\\(\\text{Gamma}(2, 0.5) + 2\\)"),
  Name = c("Group mean retention",
           "Concentration parameter for group \\(g\\)",
           "Hyperprior on \\(\\mu_g\\)",
           "Hyperprior on \\(\\kappa_g\\)"),
  Meaning = c("Average retention probability for group \\(g\\); the main parameter for cross-group comparisons",
              "Controls how tightly institutions within group \\(g\\) cluster around \\(\\mu_g\\)",
              "Encodes prior belief that group means are around 0.7 (computed average)",
              "Weakly informative prior; the \\(+2\\) shift keeps \\(\\kappa_g \\geq 2\\) to avoid U-shaped Betas"),
  Source = c("Unknown parameter, estimated by JAGS",
             "Unknown parameter, estimated by JAGS",
             "Fixed by us in the JAGS model",
             "Fixed by us in the JAGS model")
)

kbl(layer3, escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                latex_options = c("striped", "hold_position"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE, background = "#34495E", color = "white")
```

where $\alpha_g = \mu_g \kappa_g$ and $\beta_g = (1 - \mu_g) \kappa_g$,

We chose the calculated average for the Beta prior with concentration
near 71.5%.

We chose Gamma with a mean of 2 to allow kappa to range broadly. We've
also seen that sometimes it is recommended to add a constant, because
Beta distributions with a value less than 1 are U-shaped, which wouldn't
make sense with retention rates.

**From lectures: 17, 18, 19**

## Prepare data for JAGS

```{r jags_data_prep}
prepare_jags_data <- function(uni_df, cohort_col, return_col) {
  # Pull out a clean dataset for this year (drop institutions with NA or zero cohort)
  d <- uni_df %>%
    select(unitid, group, cohort = all_of(cohort_col), retained = all_of(return_col)) %>%
    filter(!is.na(cohort), !is.na(retained), cohort > 0)
  
  list(
    k          = d$retained,
    n          = d$cohort,
    group_idx  = as.integer(d$group),
    n_inst     = nrow(d),
    n_groups   = length(levels(uni_df$group)),
    institutions = d  # Keep this around for later mapping back
  )
}

jags_data_22 <- prepare_jags_data(uni, "cohort21", "return_cohort22")
jags_data_23 <- prepare_jags_data(uni, "cohort22", "return_cohort23")
jags_data_24 <- prepare_jags_data(uni, "cohort23", "return_cohort24")
```

## JAGS model

```{r jags_hier_model}
hierarchical_model <- "model {
  # Layer 1: Likelihood
  for (i in 1:n_inst) {
    k[i] ~ dbin(theta[i], n[i])
    theta[i] ~ dbeta(alpha[group_idx[i]], beta[group_idx[i]])
  }

  # Layer 2/3: Group-level hyperparameters
  for (g in 1:n_groups) {
    alpha[g] <- mu[g] * kappa[g]
    beta[g]  <- (1 - mu[g]) * kappa[g]
    
    mu[g]        ~ dbeta(7, 3)              # Hyperprior on group mean
    kappa[g]    <- kappa_raw[g] + 2          # Shift to avoid U-shaped Betas
    kappa_raw[g] ~ dgamma(2, 0.5)
  }
}"
```

**From lectures: 8, 16**

## Compile and fit

```{r}
fit_hierarchical <- function(jags_data, model_string, 
                             n_burn = 2000, n_iter = 10000, n_chains = 5) {
  # Strip the bookkeeping element before passing to JAGS
  data_for_jags <- jags_data[c("k", "n", "group_idx", "n_inst", "n_groups")]
  
  model <- jags.model(textConnection(model_string),
                      data    = data_for_jags,
                      n.chains = n_chains,
                      quiet   = TRUE)
  
  update(model, n.iter = n_burn, progress.bar = "none")  # burn-in
  
  coda.samples(model,
               variable.names = c("theta", "mu", "kappa"),
               n.iter         = n_iter,
               progress.bar   = "none")
}

posterior_22 <- fit_hierarchical(jags_data_22, hierarchical_model)
posterior_23 <- fit_hierarchical(jags_data_23, hierarchical_model)
posterior_24 <- fit_hierarchical(jags_data_24, hierarchical_model)
```

## MCMC Diagnostics

```{r mcmc_diag}
# Pull out just the mu samples
mu_samples_24 <- posterior_24[, grep("^mu\\[", varnames(posterior_24))]

# Trace plots ("hairy caterpillar" check from L16)
mcmc_trace(mu_samples_24) + theme_minimal()

# Effective sample size (should be in the thousands per parameter)
effectiveSize(mu_samples_24)

# Gelman-Rubin shrink factor (should be very close to 1.0, definitely under 1.1)
gelman.diag(mu_samples_24, multivariate = FALSE)
```

**From lectures: 16**

-- Create MCMC

## Group-level posterior summaries

```{r group_post_summ}
extract_mu_summary <- function(posterior_sample, group_levels, year_label) {
  mu_draws <- as.matrix(posterior_sample[, grep("^mu\\[", varnames(posterior_sample))])
  colnames(mu_draws) <- group_levels
  
  data.frame(
    year  = year_label,
    group = group_levels,
    mean  = apply(mu_draws, 2, mean),
    lower = apply(mu_draws, 2, quantile, 0.025),
    upper = apply(mu_draws, 2, quantile, 0.975)
  )
}

mu_summary_all <- bind_rows(
  extract_mu_summary(posterior_22, levels(uni$group), "2022"),
  extract_mu_summary(posterior_23, levels(uni$group), "2023"),
  extract_mu_summary(posterior_24, levels(uni$group), "2024")
)

mu_summary_all %>%
  kbl(digits = 4, caption = "Group-level posterior means and 95% credible intervals by year") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE, background = "#34495E", color = "white")
```

The hierarchical model's group-level posterior means are pretty
consistent across the three cohort years, with each group's posterior
mean varying by less than 0.02 across years. This stability provides
strong evidence that the observed group differences are not artifacts of
any single year. Across all three years, HSI-Commuter institutions have
the highest posterior mean retention (μ around 0.73), while
NonHSI-Commuter institutions have the lowest (μ around 0.66). To note:
the credible intervals for NonHSI-Commuter institutions are notably
wider than the others, reflecting the smaller number of institutions in
this group and the corresponding posterior uncertainty.

## Forest plot for cross-year comparison

```{r plot_cross_comp}
ggplotly(ggplot(mu_summary_all, aes(x = mean, y = group, color = year)) +
  geom_point(position = position_dodge(width = 0.5), size = 3) +
  geom_errorbarh(aes(xmin = lower, xmax = upper),
                 position = position_dodge(width = 0.5), height = 0.2) +
  labs(x = "μ_g (group mean retention probability)",
       y = NULL,
       color = "Cohort year",
       title = "Group-level retention posteriors across three cohorts") +
  theme_minimal() +
  theme(legend.position = "bottom"))
```

The forest plot displays group-level posterior means with 95% credible
intervals across the three cohort years. The near-perfect overlap of
intervals within each group confirms the yearly stability seen in the
summary table. The HSI-Commuter group consistently shows the highest
mean retention with the tightest credible interval, reflecting the
abundance of institutions in this category. By contrast, the
NonHSI-Commuter group shows wide intervals across all years.

## Institution-level posteriors / shrinkage

```{r ins_lvl_shrink_24}
#| label: fig-side-by-side
#| fig-cap: "Posterior densities (left) and institution-level shrinkage (right), 2024 cohort"
#| layout-ncol: 2
#| column: page
#| fig-width: 7
#| fig-height: 5
#| out-width: "100%"

theta_samples_24 <- posterior_24[, grep("^theta\\[", varnames(posterior_24))]
theta_draws <- as.matrix(theta_samples_24)

inst_24 <- jags_data_24$institutions %>%
  mutate(
    observed     = retained / cohort,
    theta_mean   = apply(theta_draws, 2, mean),
    theta_lower  = apply(theta_draws, 2, quantile, 0.025),
    theta_upper  = apply(theta_draws, 2, quantile, 0.975)
  ) %>%
  arrange(group, theta_mean) %>%
  mutate(order_id = row_number())

# Plot 1: Ridgeline
mu_draws_24 <- as.matrix(posterior_24[, grep("^mu\\[", varnames(posterior_24))])
colnames(mu_draws_24) <- levels(uni$group)

mu_long <- as.data.frame(mu_draws_24) %>%
  pivot_longer(everything(), names_to = "group", values_to = "mu")

ridge_plot <- ggplotly(ggplot(mu_long, aes(x = mu, y = group, fill = group)) +
  geom_density_ridges(alpha = 0.7) +
  labs(x = expression(mu[g]), y = NULL,
       title = "Group-level posteriors") +
  theme_minimal() +
  theme(legend.position = "none"))

# Plot 2: Shrinkage
shrink_plot <- ggplotly(ggplot(inst_24, aes(x = order_id, color = group)) +
  geom_point(aes(y = observed), shape = 4, size = 2, alpha = 0.6) +
  geom_point(aes(y = theta_mean), size = 2) +
  geom_errorbar(aes(ymin = theta_lower, ymax = theta_upper), width = 0.2) +
  labs(x = "Institution",
       y = "Retention probability",
       title = "Institution-level posteriors",
       color = "Group") +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        legend.position = "bottom"))

subplot(ridge_plot, shrink_plot, nrows = 1, margin = 0.05)
```

The shrinkage plot displays each institution's observed retention rate
(X markers) alongside its posterior mean and 95% credible interval (dots
and bars) for the 2024 cohort. Institutions with smaller cohorts show
greater shrinkage toward their group mean, while institutions with
larger cohorts retain estimates close to their observed rates. This is
the partial pooling property of the hierarchical model in action: rather
than treating each institution as fully independent (which would produce
noisy estimates for small schools) or pooling all institutions together
(which would mask real institutional differences), the model uses
information from peer institutions within the same group to inform each
individual estimate.

## Posterior probability statements from the hierarchical model

```{r hier_post_prob}
mu_draws_24 <- as.matrix(posterior_24[, grep("^mu\\[", varnames(posterior_24))])
colnames(mu_draws_24) <- levels(uni$group)

probs <- data.frame(
  Statement = c(
    "P(HSI_Commuter > HSI_Residential)",
    "P(HSI_Commuter > NonHSI_Commuter)",
    "P(HSI_Commuter > NonHSI_Residential)",
    "P(HSI_Commuter > 0.70)"
  ),
  Probability = c(
    1 - (mean(mu_draws_24[, "HSI_Commuter"] < mu_draws_24[, "HSI_Residential"])),
    1 - (mean(mu_draws_24[, "HSI_Commuter"] < mu_draws_24[, "NonHSI_Commuter"])),
    1 - (mean(mu_draws_24[, "HSI_Commuter"] < mu_draws_24[, "NonHSI_Residential"])),
    1 - (mean(mu_draws_24[, "HSI_Commuter"] < 0.70))
  )
)

probs %>%
  kbl(digits = 4, caption = "Selected posterior probabilities (2024 cohort)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE, background = "#34495E", color = "white")
```

The posterior probabilities quantify the strength of evidence for each
comparison. There is an 84% posterior probability that HSI-Commuter
institutions have a higher true retention rate than HSI-Residential
institutions, and an 83% posterior probability that HSI-Commuter
retention exceeds 70%. The probability that HSI-Commuter retention
exceeds NonHSI-Residential retention is more modest at 62%, suggesting
these two groups are most similar in our analysis. These probabilities
are direct outputs of the Bayesian framework and provide a more
intuitive interpretation than frequentist p-values.

## Hierarchical Reflection

Our main result is counterintuitive in a good way: HSI-Commuter
institutions in Texas have the highest first-year retention rates, not
the lowest. This challenges the narrative that schools serving Hispanic
and commuter students are necessarily underperforming. Three years of
independent hierarchical models gave us nearly identical results, which
is rare in education research and gives us strong confidence in the
conclusions. Lastly, our hierarchical model can't tell us why
HSI-Commuter schools are doing better. We have correlation, not
causation. It could be Texas-specific policy, it could be selection
effects, it could be measurement artifacts. This could be a question for
follow-up research.

# Part 3: Model Comparison

Compare conjugate pooled posteriors against hierarchical group-level
posteriors for the same four groups. The hierarchical intervals will be
wider for small groups (appropriate acknowledgment of institutional
heterogeneity) and may shift means slightly toward the overall data
pattern.

\*Make sure to include credible intervals from conjugate and
hierarchical models

-- Maybe try a multinomial logistic regression to compare all the groups
adds rigor, rather than doing all the groups independently.

# Conclusion:

Points to hit:

-   Key findings on HSI vs non-HSI and commuter vs residential retention
-   Specifically what the posteriors say about institutions like UHD
-   The value added by the hierarchical model (shrinkage, institutional
    heterogeneity)
-   Limitations: single year, institution-level aggregation ignores
    student-level predictors, etc.
-   Future directions: multi-year trend, student-level data, additional
    covariates

**From lectures: 5** - Has good language on interpretations

# References

National Center for Education Statistics. (2024). *Integrated
Postsecondary Education Data System (IPEDS)*. U.S. Department of
Education, Institute of Education Sciences. <https://nces.ed.gov/ipeds/>

Andriano, B. C. (2020). *Perceptions of commuter students and faculty: A
mixed methods study on commuter student retention* \[Doctoral
dissertation, Lindenwood University\]. Digital Commons \@ Lindenwood
University. <https://digitalcommons.lindenwood.edu/dissertations/187/>

Postsecondary National Policy Institute. (2021). *Hispanic-serving
institutions* \[Policy primer\].
<https://pnpi.org/wp-content/uploads/2021/09/HSI-Primer_Updated_September2021.pdf>

Every Learner Everywhere. (2025). *A profile of Hispanic-serving
institutions*.
<https://www.everylearnereverywhere.org/resources/a-profile-of-hispanic-serving-institutions/>
