---
title: "Bayesian Estimation of First-Year Retention Rates: A Comparative Analysis of Commuter and Hispanic-Serving Institutions in Texas"
author:   
  - Eugene Kazi
  - Shikyna William
  - Rehana Hajee
date: "`r Sys.Date()`"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    code-fold: true
    theme: flatly
    highlight-style: tango
    include-in-header:
      - text: |
          <style>
          .quarto-title-block .quarto-title-banner,
          .quarto-title-block .quarto-title {
            text-align: center;
          }
          .quarto-title-meta > div {
            text-align: center;
          }
          </style>
  #pdf:
    #toc: true
    #number-sections: true
execute:
  warning: false
  message: false
  echo: true
editor: 
  markdown: 
    wrap: 72
---

# Introduction, Problem, Purpose

This section should include the data and problem descriptions, your
background research and information on related studies, information as
to how your data were collected (with proper referencing) , and the
hypotheses or questions you want answered using Statistics. A statement
as to why your project is important or relevant should also be included.

## Setup & data preparation

```{r, include=FALSE}
library(tidyverse)
library(rjags)
library(coda)
library(bayesplot)
library(knitr)
library(kableExtra)
library(plotly)
```

```{r, message=FALSE, warning=FALSE}
y22 <- read_csv("Bayes CommuterHSI Data/2223.csv")
y23 <- read_csv("Bayes CommuterHSI Data/2324.csv")
y24 <- read_csv("Bayes CommuterHSI Data/2425.csv")
hsi22 <- read_csv("Bayes CommuterHSI Data/HSI22.csv")
hsi23 <- read_csv("Bayes CommuterHSI Data/HSI23.csv")
hsi24 <- read_csv("Bayes CommuterHSI Data/HSI24.csv")

y22 <- y22 %>%
  select(-starts_with("CUSTOMCGIDS"), -'institution name...4', -'HD2022.State abbreviation', -'IC2022_PCCAMPUSES.Sector of institution') %>% distinct()
y23 <- y23 %>%
  select(-starts_with("CUSTOMCGIDS"), -'institution name...4', -'HD2023.State abbreviation', -'IC2023_PCCAMPUSES.Sector of institution') %>% distinct()
y24 <- y24 %>%
  select(-starts_with("CUSTOMCGIDS"), -'institution name...4', -'HD2024.State abbreviation', -'XROOMCAP', -'XRET_PCF', -'XRET_NMF', -'XRRFTCTA', -'HD2024.Sector of institution') %>% distinct()

hsi22 <- hsi22 %>% 
  select(-'IDX_EF', -'EF2022A.Level of student')
hsi23 <- hsi23 %>% 
  select(-'IDX_EF', -'EF2023A.Level of student')
hsi24 <- hsi24 %>% 
  select(-'IDX_EF', -'EF2024A.Level of student') %>% rename('hisp_pop' = 'EF2024A.Hispanic total', 'tot_pop' = 'EF2024A.Grand total')

y22 <- y22 %>% rename('name' = 'institution name...2', 'housing' = 'IC2022.Institution provide on-campus housing', 'dorm_cap' = 'IC2022.Total dormitory capacity', 'cohort21' = 'EF2022D_RV.Full-time adjusted fall 2021 cohort', 'return_cohort22' = 'EF2022D_RV.Students from the full-time adjusted fall 2021 cohort enrolled in fall 2022', 'ret_rate22' = 'EF2022D_RV.Full-time retention rate, 2022')

y23 <- y23 %>% rename('name' = 'institution name...2', 'housing' = 'IC2023.Institution provide nstitutionally-controlled housing (on-campus and/or off-campus)', 'dorm_cap' = 'IC2023.Housing capacity', 'cohort22' = 'EF2023D_RV.Full-time adjusted fall 2022 cohort', 'return_cohort23' = 'EF2023D_RV.Students from the full-time adjusted fall 2022 cohort enrolled in fall 2023', 'ret_rate23' = 'EF2023D_RV.Full-time retention rate, 2023')

y24 <- y24 %>% rename('name' = 'institution name...2', 'housing' = 'COST1_2024.Institution provide Institutionally-controlled housing (on-campus and/or off-campus)', 'dorm_cap' = 'COST1_2024.Housing capacity', 'cohort23' = 'EF2024D.Full-time adjusted fall 2023 cohort', 'return_cohort24' = 'EF2024D.Students from the full-time adjusted fall 2023 cohort enrolled in fall 2024', 'ret_rate24' = 'EF2024D.Full-time retention rate, 2024') %>% filter(cohort23 > 0)

# Classify institutions: HSI & Commuter Status
y24 <- left_join(y24, select(hsi24, unitid, tot_pop, hisp_pop), by='unitid') %>% mutate('hsi' = (hisp_pop / tot_pop) > .25, 'commuter' = (dorm_cap / tot_pop) < .25)
y24$commuter[y24$housing == "No"] <- T

y24 <- y24 %>% select(-'housing', -'dorm_cap', -'hisp_pop', -'year')

uni <- y24 %>% left_join(select(y23, 'unitid', 'cohort22', 'return_cohort23', 'ret_rate23'), by='unitid') %>% left_join(select(y22, 'unitid', 'cohort21', 'return_cohort22', 'ret_rate22'), by='unitid')

uni <- uni %>% select("unitid", "name", "tot_pop", "hsi", "commuter", "cohort23", "return_cohort24", "ret_rate24", "cohort22", "return_cohort23", "ret_rate23", "cohort21", "return_cohort22", "ret_rate22")
uni <- uni %>% filter(unitid != "229355")
```

In the final table, we have `r ncol(uni)` variables. This includes; the
school's ID, name, population, commuter and HSI flags, and the cohort
and retention rate information for the years 2022, 2023, and 2024.

## Descriptive Summary

```{r}
uni %>% select(tot_pop, hsi, commuter, ret_rate24, ret_rate23, ret_rate22) %>% summarise(
  n_schools = nrow(uni),
  hsi_proportion = round(sum(uni$hsi) / nrow(uni),  2),
  commuter_proportion = round(sum(uni$commuter) / nrow(uni), 2),
  avg_ret_rate24 = mean(ret_rate24),
  avg_ret_rate23 = mean(ret_rate23),
  avg_ret_rate22 = mean(ret_rate22),
) %>% pivot_longer(everything(), names_to = "Statistic", values_to = "Value") %>%
  kbl(digits = 3, caption = "Summary of Texas public 4-year institutions") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    position = "center"
  )
```

# Part 1: Conjugate Beta-Binomial Analysis (base model)

The baseline model pools all students within each of the four groups:

$$X_g \mid n_g, \theta_g \sim \text{Binomial}(n_g, \theta_g)$$

```{r}
# Likelihood
likelihood_p1 <- data.frame(
  Symbol = c("\\(X_g\\)", "\\(n_g\\)", "\\(\\theta_g\\)", "\\(\\text{Bin}(n, \\theta)\\)"),
  Name = c("Pooled retained count (random variable)",
           "Pooled cohort size",
           "True retention probability for group \\(g\\)",
           "Binomial distribution"),
  Meaning = c("Total number of students across all institutions in group \\(g\\) who returned for their second year",
              "Total number of full-time first-time students across all institutions in group \\(g\\)",
              "Unknown true probability that a student in group \\(g\\) returns for their second year",
              "Out of \\(n\\) independent trials, each succeeding with probability \\(\\theta\\)"),
  Source = c("Observed data, summed within group",
             "Observed data, summed within group",
             "Unknown parameter, estimated analytically",
             "Distribution shape")
)

kbl(likelihood_p1, escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                latex_options = c("striped", "hold_position"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE, background = "#34495E", color = "white")
```

$$\theta_g \sim \text{Beta}(\alpha_0, \beta_0)$$

```{r}
# Prior
prior_p1 <- data.frame(
  Symbol = c("\\(\\theta_g\\)", "\\(\\alpha\\)", "\\(\\beta\\)", "\\(\\text{Beta}(\\alpha, \\beta)\\)"),
  Name = c("True retention probability for group \\(g\\)",
           "Prior shape parameter (alpha)",
           "Prior shape parameter (beta)",
           "Beta distribution"),
  Meaning = c("Same as in the likelihood",
              "Interpretable as the number of \"prior successes\"",
              "Interpretable as the number of \"prior failures\"",
              "Flexible distribution on \\([0, 1]\\) defined by two shape values"),
  Source = c("Unknown parameter",
             "Fixed by us (varies by prior choice: 1, 2, or 7)",
             "Fixed by us (varies by prior choice: 1, 2, or 3)",
             "Distribution shape")
)

kbl(prior_p1, escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                latex_options = c("striped", "hold_position"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE, background = "#34495E", color = "white")
```

The conjugate posterior would be:

$$\theta_g \mid x_g \sim \text{Beta}(\alpha_0 + x_g, \beta_0 + n_g - x_g)$$

```{r}
# Posterior
posterior_p1 <- data.frame(
  Symbol = c("\\(\\theta_g \\mid x_g\\)", "\\(\\alpha + x_g\\)", "\\(\\beta + n_g - x_g\\)"),
  Name = c("Posterior distribution of \\(\\theta_g\\) given the data",
           "Updated alpha (posterior shape parameter 1)",
           "Updated beta (posterior shape parameter 2)"),
  Meaning = c("Our updated belief about \\(\\theta_g\\) after observing \\(x_g\\) retained students out of \\(n_g\\)",
              "Prior alpha plus observed successes",
              "Prior beta plus observed failures"),
  Source = c("Conjugate result (closed-form Beta)",
             "Computed from data and prior",
             "Computed from data and prior")
)

kbl(posterior_p1, escape = FALSE, caption = "Posterior (conjugate update)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                latex_options = c("striped", "hold_position"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE, background = "#34495E", color = "white")
```

**From Lectures: 7, 9**

## Pool the data into groups

We would group each and find retained cohort by year using summarize.

## Test three prior choices for sensitivity analysis

We can test;

Uninformative: Beta (1, 1) - Completely flat prior

```{r, echo=F}
plot_ly(x = seq(0, 1, length.out = 100), y = dbeta(seq(0, 1, length.out = 100), 1, 1), type = 'scatter', mode = 'lines', fill = 'tozeroy') %>%
  layout(
    title = paste0("Beta Distribution (Alpha =", 1, ", Beta =", 1, ")"),
    xaxis = list(title = "x"),
    yaxis = list(title = "Density")
  )
```

\
Weakly Informative: Beta (2, 2) - Slight pull to 0.5

```{r, echo=F}
plot_ly(x = seq(0, 1, length.out = 100), y = dbeta(seq(0, 1, length.out = 100), 2, 2), type = 'scatter', mode = 'lines', fill = 'tozeroy') %>%
  layout(
    title = paste0("Beta Distribution (Alpha =", 2, ", Beta =", 2, ")"),
    xaxis = list(title = "x"),
    yaxis = list(title = "Density")
  )
```

Informative: Beta (7, 3) - Centered near our calculated 71.5% average
retention rate.

```{r, echo=F}
plot_ly(x = seq(0, 1, length.out = 100), y = dbeta(seq(0, 1, length.out = 100), 7, 3), type = 'scatter', mode = 'lines', fill = 'tozeroy') %>%
  layout(
    title = paste0("Beta Distribution (Alpha =", 7, ", Beta =", 3, ")"),
    xaxis = list(title = "x"),
    yaxis = list(title = "Density")
  )
```

\*Make sure to include credible intervals

-- Maybe consider using posterior of previous years as prior for
following years.

-- Use posterior predictive to check adequacy of our model.

**From lectures: 4, 9, 10**

## Compute posteriors under each prior

## Plot posterior densities (main prior)

## Posterior probability statements

Such as probability HSI-Commuter group has lower retention than
residential groups

# Part 2: Hierarchical Beta-Binomial Model (Main Analysis)

Because retention is a count of successes out of a fixed cohort, we use
the Binomial likelihood with a conjugate Beta prior rather than the
Normal-Normal hierarchy from class (l18).

The conjugate model we used assumes all the institutions within a group
share the same retention probability, which is kind of unrealistic. The
hierarchical model allows each institution (i) have its own (theta_i),
but institutions within the same group (g) share a common Beta
distribution:

$$\text{Layer 1 (likelihood):} \quad X_{ig} \mid n_{ig}, \theta_{ig} \sim \text{Binomial}(n_{ig}, \theta_{ig})$$

```{r}
# Layer 1: Likelihood
layer1 <- data.frame(
  Symbol = c("\\(X_{ig}\\)", "\\(n_{ig}\\)", "\\(\\theta_{ig}\\)", "\\(\\text{Bin}(n, \\theta)\\)"),
  Name = c("Retained count (random variable)",
           "Cohort size",
           "True retention probability",
           "Binomial distribution"),
  Meaning = c("Number of students from institution \\(i\\) in group \\(g\\) who returned for their second year",
              "Total full-time first-time students at institution \\(i\\) in group \\(g\\) who entered the previous fall",
              "Unknown true probability that a student at institution \\(i\\) in group \\(g\\) returns for year two",
              "Out of \\(n\\) independent trials, each succeeding with probability \\(\\theta\\)"),
  Source = c("Observed data (`retained` column)",
             "Observed data (`cohort_n` column)",
             "Unknown parameter, estimated by JAGS",
             "Distribution shape")
)

kbl(layer1, escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                latex_options = c("striped", "hold_position"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE, background = "#34495E", color = "white")
```

$$\text{Layer 2 (prior):} \quad \theta_{ig} \mid \alpha_{g}, \beta_g \sim \text{Beta}(\alpha_g, \beta_g)$$

```{r}
# Layer 2: Prior
layer2 <- data.frame(
  Symbol = c("\\(\\theta_{ig}\\)", "\\(\\alpha_g\\)", "\\(\\beta_g\\)", "\\(\\text{Beta}(\\alpha, \\beta)\\)"),
  Name = c("True retention probability",
           "Beta shape parameter (alpha) for group \\(g\\)",
           "Beta shape parameter (beta) for group \\(g\\)",
           "Beta distribution"),
  Meaning = c("Same as Layer 1",
              "Prior successes parameter for the group-level Beta",
              "Prior failures parameter for the group-level Beta",
              "Flexible distribution on \\([0, 1]\\) defined by two shape values"),
  Source = c("Unknown parameter",
             "Computed: \\(\\alpha_g = \\mu_g \\cdot \\kappa_g\\)",
             "Computed: \\(\\beta_g = (1 - \\mu_g) \\cdot \\kappa_g\\)",
             "Distribution shape")
)

kbl(layer2, escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                latex_options = c("striped", "hold_position"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE, background = "#34495E", color = "white")
```

$$\text{Layer 3 (hyperprior):} \quad \mu_g \sim \text{Beta}(7, 3), \quad \kappa_g \sim \text{Gamma}(2, 0.5) + 2$$

```{r}
# Layer 3: Hyperprior
layer3 <- data.frame(
  Symbol = c("\\(\\mu_g\\)", "\\(\\kappa_g\\)", "\\(\\text{Beta}(7, 3)\\)", "\\(\\text{Gamma}(2, 0.5) + 2\\)"),
  Name = c("Group mean retention",
           "Concentration parameter for group \\(g\\)",
           "Hyperprior on \\(\\mu_g\\)",
           "Hyperprior on \\(\\kappa_g\\)"),
  Meaning = c("Average retention probability for group \\(g\\); the main parameter for cross-group comparisons",
              "Controls how tightly institutions within group \\(g\\) cluster around \\(\\mu_g\\)",
              "Encodes prior belief that group means are around 0.7 (computed average)",
              "Weakly informative prior; the \\(+2\\) shift keeps \\(\\kappa_g \\geq 2\\) to avoid U-shaped Betas"),
  Source = c("Unknown parameter, estimated by JAGS",
             "Unknown parameter, estimated by JAGS",
             "Fixed by us in the JAGS model",
             "Fixed by us in the JAGS model")
)

kbl(layer3, escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                latex_options = c("striped", "hold_position"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE, background = "#34495E", color = "white")
```

where $\alpha_g = \mu_g \kappa_g$ and $\beta_g = (1 - \mu_g) \kappa_g$,

We chose the calculated average for the Beta prior with concentration
near 71.5%.

We chose Gamma with a mean of 4 to allow kappa to range broadly. I also
seen that sometimes it is recommended to add a constant, because Beta
distributions with a value less than 1 are U-shaped, which wouldn't make
sense with retention rates.

**From lectures: 17, 18, 19**

## Prepare data for JAGS

## JAGS model

**From lectures: 8, 16**

## Compile and fit

```{r}
# model <- 

# update()

# posterior_sample
```

## If we're doing MCMC, we do that now

**From lectures: 16**

-- Create MCMC

## Group-level posterior summaries

\*Make sure to include credible intervals for each group's
hyperparameter mean using quantile(mu_samples, c(0.025, 0.975))

-- Plot & compare credible intervals by year for each group

## Institution-level posteriors / shrinkage

\*Make sure to include credible intervals. This is where the smaller
schools will have their intervals pulled toward the group mean more than
larger schools.

## Posterior probability statements from the hierarchical model

# Part 3: Model Comparison

Compare conjugate pooled posteriors against hierarchical group-level
posteriors for the same four groups. The hierarchical intervals will be
wider for small groups (appropriate acknowledgment of institutional
heterogeneity) and may shift means slightly toward the overall data
pattern.

\*Make sure to include credible intervals from conjugate and
hierarchical models

-- Maybe try a multinomial logistic regression to compare all the groups
adds rigor, rather than doing all the groups independently.

## Bayesian Anova

**From lectures: 17, 19, 20**

# Conclusion:

Points to hit:

-   Key findings on HSI vs non-HSI and commuter vs residential retention
-   Specifically what the posteriors say about institutions like UHD
-   The value added by the hierarchical model (shrinkage, institutional
    heterogeneity)
-   Limitations: single year, institution-level aggregation ignores
    student-level predictors, etc.
-   Future directions: multi-year trend, student-level data, additional
    covariates

**From lectures: 5** - Has good language on interpretations

# References

IPEDS
